BOLD Parameter Estimation Using Particle Filters

### 2010-07-19 doc/thesis/background.tex
More clean up.

\begin{figure}
\caption{Hemodynamic Response Function todo}
\label{fig:HRF}
\end{figure}

\subsection{Hierarchical Linear Models}
As mentioned previously and discussed extensively
in \cite{Friston2002} and \cite{Hoffman1997}, hierarchical models may be 
applied to account for systematic differences between subjects. For instance, if the
study happens to be a mix between young and old, incorporating
that into the model is wise, regardless if that is the purpose of the test. 
The reason to do this is to account for additional variance that may not
be present in similar studies. The Hierarchical form used by
\cite{Friston2002} is shown in \autoref{eq:Hiearchical}.

@@

The Empirical Bayes algorithm is used in both in both \cite{Friston2002} 
and \cite{Hoffman1997}. As a consequence, point estimators 
are used for each $\theta$, rather than the full distributions.

\subsection{Conclusion}
\label{sec:BackgroundConclusion}
In all, the GLM is extremely useful for determining linear 
dependence of a set of regressors on the output. Unfortunately, as discussed in
\autoref{sec:BOLD Analysis} there are significant nonlinearities 
that almost certainly cause false negatives in the Statistical Parametric
Maps. Unfortunately nonlinear analyses have only recently become feasible,
so the scale of the problem is still unknown. The problem is 
highlighted by the relatively common scenario where no significant
activation can be found in a single run \cite{Riera2004}
\cite{Johnston2008}.

The static nature of the  linear model also limits its inference power. 
Besides not permitting HRF differences between patients, there is no
reasonable way to incorporate other forms of physiological
data. Combined FMRI CBF or CBV imaging methods are rapidly getting better,
as seen in \cite{Chan2009}. These techniques could shed light on
neural activation by providing extra measurements, yet a 
physiologically reasonable model is necessary for this.
In reverse, activation detection methods also don't have the ability 
to identify pathologies based on state variables or parameters. For
example decreased compliance of
blood vessels could indicate, or even cause, a neurological condition that 
is not easily seen in other imaging modalities. Thus, the benefits
of physiologically meaningful models are manifold.

@@

Although both these methods can be highly effective, they have the downside of
requiring very high computation time. In this case of the BOLD model,
each time the energy or fitness needs to be calculated, a large number of cycles
must be spent re-simulating the BOLD model for the set of parameters. As we'll
discuss in \autoref{sec:Particle Filters}, the Particle Filter method is able
to circumvent this re-calculation to some degree.

@@

\subsection{Hybrid Methods}
In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used, as described by
\cite{Ozaki1994}. \cite{Ozaki1994} uses a similar construction to a 
Kalman filter, to break the time series into a series of innovations,
for which Maximum Likelihood was performed. While this is probably the most
likely solution to give the correct output, there are a few problems. First, every
step in parameter space requires a recalculation of all the state variables. With
two or three parameters this is fine, more than that, and calculations can become
intractable. Additionally, there is no way to ensure a global minimum is reached.
Although this is also true of the methods in \autoref{sec:Nonlinear Least Squares},
those methods contained random elements designed to overcome this issue.
[results?]

In \cite{Johnston2008}, a hybrid particle filter/gradient
descent algorithm was used to simultaneously derive the static and dynamic 
parameters, (classically known as parameters and state variables, respectively).
Essentially a particle filter is used to calculate the state variables at each
time; then the estimated distribution of the particles is used to find
the most likely set of parameters that would give that distribution of state variables.
This process is repeated until the parameters converge. Interestingly
\cite{Johnston2008} comes to a very different set of parameter estimates as compared
to the original \cite{Friston2000} estimates (\autoref{tab:Params}). In fact the results 
are significantly different from virtually every other study. The most obvious discrepancy
is the significantly longer time constants, $\tau_f$, $\tau_s$ and $\tau_0$. 
While of course this could be poor convergence of the algorithm, there is another other possibility.
Unlike all the other methods mentioned, excepting the methods in \autoref{sec:Least Squares},
the algorithm described in \cite{Johnston2008} does not depend on prior distributions.
It is possible then that the bias toward the prior in other methods screwed the results. 
While \cite{Johnston2008} is certainly in the minority; further exhaustive studies 
of the parameters, using unbiased techniques may be called for.

[graph of friston parameters vs. johnston]

[might want to mention dynamic causal modeling]

>>>>

\begin{figure}
\centering
\includegraphics[scale=.7]{images/HRF}
\caption{Canonical Hemodynamic Response Function}
\label{fig:HRF}
\end{figure}

\subsection{Hierarchical Linear Models}
As mentioned previously and discussed extensively
in \cite{Friston2002} and \cite{Hofmann1997}, hierarchical models may be 
applied to account for systematic differences between subjects. For instance, if the
study happens to be a mix between young and old, incorporating
that into the model is wise, regardless if that is the purpose of the test. 
The reason to do this is to account for additional variance that may not
be present in similar studies. The Hierarchical form used by
\cite{Friston2002} is shown in \autoref{eq:Hierarchical}.

@@

The Empirical Bayes algorithm is used in both in both \cite{Friston2002} 
and \cite{Hofmann1997}. As a consequence, point estimators 
are used for each $\theta$, rather than the full distributions.

\subsection{Discussion}
\label{sec:BackgroundConclusion}
In all, the GLM is extremely useful for determining linear 
dependence of a set of regressors on the output. Unfortunately, as discussed in
\autoref{sec:BOLD Analysis} there are significant nonlinearities 
that almost certainly cause false negatives in the Statistical Parametric
Maps. Unfortunately nonlinear analyses have only recently become feasible,
so the scale of the problem is still unknown. The problem is 
highlighted by the relatively common scenario where no significant
activation can be found in a single run \cite{Riera2004}
\cite{Johnston2008}.

The static nature of the  linear model also limits its inference power. 
Besides not permitting HRF differences between patients, there is no
reasonable way to incorporate other forms of physiological
data. Combined FMRI CBF or CBV imaging methods are rapidly getting better,
as seen in \cite{Chen2009}. These techniques could shed light on
neural activation by providing extra measurements, yet a 
physiologically reasonable model is necessary for this.
In reverse, activation detection methods also don't have the ability 
to identify pathologies based on state variables or parameters. For
example decreased compliance of
blood vessels could indicate, or even cause, a neurological condition that 
is not easily seen in other imaging modalities. Thus, the benefits
of physiologically meaningful models are manifold.

@@

Although both these methods can be highly effective, they have the downside of
requiring very high computation time. In this case of the BOLD model,
each time the energy or fitness needs to be calculated, a large number of cycles
must be spent re-simulating the BOLD model for the set of parameters. As I'll
discuss in \autoref{sec:Particle Filter}, the Particle Filter method is able
to circumvent this re-calculation to some degree.

@@

\subsection{Hybrid Methods}
In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used, as described by
\cite{Ozaki1994}. \cite{Ozaki1994} uses a similar construction to a 
Kalman filter, to break the time series into a series of innovations,
for which Maximum Likelihood was performed. While this is probably the most
likely solution to give the correct output, there are a few problems. First, every
step in parameter space requires a recalculation of all the state variables. With
two or three parameters this is fine, more than that, and calculations can become
intractable. Additionally, there is no way to ensure a global minimum is reached.
Although this is also true of the methods in \autoref{sec:Nonlinear Least Squares},
those methods contained random elements designed to overcome this issue.

In \cite{Johnston2008}, a hybrid particle filter/gradient
descent algorithm was used to simultaneously derive the static and dynamic 
parameters, (classically known as parameters and state variables, respectively).
Essentially a particle filter is used to calculate the state variables at each
time; then the estimated distribution of the particles is used to find
the most likely set of parameters that would give that distribution of state variables.
This process is repeated until the parameters converge. Interestingly
\cite{Johnston2008} comes to a very different set of parameter estimates as compared
to the original \cite{Friston2000} estimates (\autoref{tab:Params}). In fact the results 
are significantly different from virtually every other study. The most obvious discrepancy
is the significantly longer time constants, $\tau_f$, $\tau_s$ and $\tau_0$. 
While of course this could be poor convergence of the algorithm, there is another other possibility.
Unlike all the other methods mentioned, excepting the methods in \autoref{sec:Nonlinear Least Squares},
the algorithm described in \cite{Johnston2008} does not depend on prior distributions.
It is possible then that the bias toward the prior in other methods screwed the results. 
While \cite{Johnston2008} is certainly in the minority; further exhaustive studies 
of the parameters, using unbiased techniques may be called for. A further 
comparison between the distributions found in \cite{Johnston2008} and 
\cite{Friston2000} will be discussed in \autoref{sec:PriorDist}, including
graphs: \autoref{fig:MeanResponseF}, \autoref{fig:MeanResponseJ}.

### 2010-07-19 doc/thesis/intro.tex
More clean up.

Regardless of the probability that CMRO2 and CBF are detached,
research into the post-stimulus undershoot has led to the creation
of much more in depth models. In \cite{Zhen2002} additional state
variables model oxygen transport, whereas \cite{Buxton2004} models
CMR02 from a higher level, and somewhat more simply; though it 
still adds several new parameters. \cite{Behzadi2005}
introduces nonlinearities into the CBF equations as a method to
explain the post-stimulus undershoot, which falls in line with a 
prolonged increase in CBF observed in \cite{Chen2009}. Similarly
\cite{Zheng2005} adds additional compartments to model the BOLD signal
that result from venous and arterial blood. 
\cite{Deneux2006} compared these various models and though it did 
not deal extensively with the 
post-stimulus undershoot, it did show incremental improvements
in quality from additional parameters (over the basic Balloon model);
though at the cost of greatly increased complexity.
Importantly,\cite{Deneux2006} did show that by 
simply adding viscoelastic terms from \cite{Buxton2004}, a slowed return 
to baseline is possible to model, without greatly increasing
complexity. Regardless, because of many of these models are extremely
complex, and the benefits have yet to be proved, in this work the simple
Balloon model will be used. Simplicity is even more important because
it is the first time a particle filter has been used to calculate the
BOLD parameters. Of course, future works could certainly benefit from
the more advanced models, especially with the addition of viscoelastic
effects.

@@

Necessary?
[Image with two different $\alpha$'s]
[image comparing the results of 10\% changes in various signals]

>>>>

Regardless of the probability that CMRO2 and CBF are detached,
research into the post-stimulus undershoot has led to the creation
of much more in depth models. In \cite{Zheng2002} additional state
variables model oxygen transport, whereas \cite{Buxton2004} models
CMR02 from a higher level, and somewhat more simply; though it 
still adds several new parameters. \cite{Behzadi2005}
introduces nonlinearities into the CBF equations as a method to
explain the post-stimulus undershoot, which falls in line with a 
prolonged increase in CBF observed in \cite{Chen2009}. Similarly
\cite{Zheng2005} adds additional compartments to model the BOLD signal
that result from venous and arterial blood. 
\cite{Deneux2006} compared these various models and though it did 
not deal extensively with the 
post-stimulus undershoot, it did show incremental improvements
in quality from additional parameters (over the basic Balloon model);
though at the cost of greatly increased complexity.
Importantly,\cite{Deneux2006} did show that by 
simply adding viscoelastic terms from \cite{Buxton2004}, a slowed return 
to baseline is possible to model, without greatly increasing
complexity. Regardless, because of many of these models are extremely
complex, and the benefits have yet to be proved, in this work the simple
Balloon model will be used. Simplicity is even more important because
it is the first time a particle filter has been used to calculate the
BOLD parameters. Of course, future works could certainly benefit from
the more advanced models, especially with the addition of viscoelastic
effects.

@@

### 2010-07-19 doc/thesis/methods.tex
More clean up.

There is an interesting anomaly with the priors found in virtually all
the works that characterized the parameters, except \cite{Johnston2008}.
The BOLD signal is universally recognized to be around $2-3\%$, maybe
reaching $5\%$ in extreme activation. Yet using the mean priors
from \cite{Friston2000}, the signal response for a $.1$ second
impulse only reaches maybe half a percent, as \autoref{fig:MeanResponse}
shows.

\begin{figure}
\centering
\label{fig:MeanResponse}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/mean_response}
\caption{Response to $.1s$ impulses with the mean parameters from \cite{Friston2000}}
\end{figure}

@@

\begin{figure}
\centering
\label{fig:MeanResponse}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/mean_response_johnston}
\caption{Response to $.1s$ impulses with the mean parameters from \cite{Johnston2008}}
\end{figure}

@@

\section{Choosing $P(y_k | x_k)$}
Choosing a representation of an unknown distribution tricky, and while
the usual choice is a Gaussian distribution, there are reasons why it
may not be the best.  Studies of the noise in FMRI typically characterise
noise as an additive noise process with Gaussian steps; a Wiener process.

@@

Because of the need to err on the side of caution I tested several weighting
functions. In addition to the Gaussian I also tested the Laplace and Cauchy
distributions, both of which have much wider tales than the Gaussian. The
benefit of the wider tailed distributions is that they don't down-weight
particles quite as fast. Additionally, the Laplace distribution has the
benefit of having a non-zero slope at the origin. This means that even
if the distribution is made overly broad, it will still distinguish between
particles that are near the origin. Results with real data compare 
the three distribution in \autoref{sec:Results Weights}.

@@

Even if $\nu_d$ is some other additive process, the difference will still be closer
to I.I.D. than a Wiener process, as the autocorrelation of the $\delta y$ shows
in \autoref{fig:QQDDelta} in \autoref{sec:Introduction Noise}. 
 All the assumptions made originally
for the particle filter still hold, and all of the parameters may be distinguished based on
the step sizes, thus it is not unreasonable to consider matching the string of step sizes
rather than string of direct readings.

>>>>

There is an interesting anomaly with the priors found in virtually all
the works that characterized the parameters, except \cite{Johnston2008}.
The BOLD signal is universally recognized to be around $2-3\%$, maybe
reaching $5\%$ in extreme activation. Yet using the mean priors
from \cite{Friston2000}, the signal response for a $.1$ second
impulse only reaches maybe half a percent, as \autoref{fig:MeanResponseF}
shows.

\begin{figure}
\centering
\label{fig:MeanResponseF}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/mean_response}
\caption{Response to $.1s$ impulses with the mean parameters from \cite{Friston2000}}
\end{figure}

@@

\begin{figure}
\centering
\label{fig:MeanResponseJ}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/mean_response_johnston}
\caption{Response to $.1s$ impulses with the mean parameters from \cite{Johnston2008}}
\end{figure}

@@

\section{Choosing $P(y_k | x_k)$}
\label{sec:Methods Weighting Function}
Choosing a representation of an unknown distribution tricky, and while
the usual choice is a Gaussian distribution, there are reasons why it
may not be the best.  Studies of the noise in FMRI typically characterise
noise as an additive noise process with Gaussian steps; a Wiener process.

@@

Because of the need to err on the side of caution I tested several weighting
functions. In addition to the Gaussian I also tested the Laplace and Cauchy
distributions, both of which have much wider tales than the Gaussian. The
benefit of the wider tailed distributions is that they don't down-weight
particles quite as fast. Additionally, the Laplace distribution has the
benefit of having a non-zero slope at the origin. This means that even
if the distribution is made overly broad, it will still distinguish between
particles that are near the origin.

@@

Even if $\nu_d$ is some other additive process, the difference will still be closer
to I.I.D. than a Wiener process, as the autocorrelation of the $\delta y$ shows
in \autoref{fig:QQDelta} in \autoref{sec:Introduction Noise}. 
 All the assumptions made originally
for the particle filter still hold, and all of the parameters may be distinguished based on
the step sizes, thus it is not unreasonable to consider matching the string of step sizes
rather than string of direct readings.
